# (PART) Demonstration {-}

# R as GIS: Demonstrations {#demo}

```{r setup, include = FALSE, cache = FALSE}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  # cache = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  cache.lazy = FALSE
)

library(here)
opts_knit$set(root.dir = here())
options(knitr.duplicate.label = "allow")
```

```{r ch01-setwd, eval = FALSE, echo = FALSE}
setwd(here())
```

```{r, include = FALSE, cache = FALSE}
#--- load packages ---#
library(data.table)
library(here)
library(rgdal)
library(stringr)
library(rgeos)
library(sf)
library(ggplot2)
library(raster)
library(modelsummary)
library(tmap)
library(future.apply)
library(lubridate)
library(fixest)
library(tidyverse)

#--- set seed ---#
set.seed(2473498)
```

## Before you start {-}

The primary objective of this chapter is to showcase the power of R as GIS through demonstrations via mock-up econometric research projects (first five demonstrations) and creating variables used in published articles (the last three demonstrations)^[Note that this lecture does not deal with spatial econometrics at all. This lecture is about spatial data processing, not spatial econometrics. [This](http://www.econ.uiuc.edu/~lab/workshop/Spatial_in_R.html) is a great resource for spatial econometrics in R.]. Each demonstration project consists of a project overview (objective, datasets used, econometric model, and GIS tasks involved) and demonstration. This is really **NOT** a place you learn the nuts and bolts of how R does spatial operations. Indeed, we intentionally do not explain all the details of how the R codes work. We reiterate that the main purpose of the demonstrations is to get you a better idea of how R can be used to process spatial data to help your research projects involving spatial datasets. Finally, note that the *mock-up* projects use extremely simple econometric models that completely lacks careful thoughts you would need in real research projects. So, don't waste your time judging the econometric models, and just focus on GIS tasks. If you are not familiar with html documents generated by `rmarkdown`, you might benefit from reading the conventions of the book in the Preface. Finally, for those who are interested in replicating the demonstrations, directions for replication are provided below. However, I would suggest focusing on the narratives for the first time around, learn the nuts and bolts of spatial operations from Chapters 2 through 5, and then come back to replicate them. 

### Target Audience {-}

The target audience of this chapter is those who are not very familiar with R as GIS. Knowledge of R certainly helps. But, I tried to write in a way that R beginners can still understand the power of R as GIS^[I welcome any suggestions to improve the reading experience of inexperienced R users.]. Do not get bogged down by all the complex-looking R codes. Just focus on the narratives and figures to get a sense of what R can do.

### Direction for replication {-}

---

**Datasets**

Running the codes in this chapter involves reading datasets from a disk. All the datasets that will be imported are available [here](https://www.dropbox.com/sh/rtbs4ji21c9uiy9/AADYpHAWhUxMittAptuq-Apaa?dl=0). In this chapter, the path to files is set relative to my own working directory (which is hidden). To run the codes without having to mess with paths to the files, follow these steps:^[I thought about using the `here` package, but I found it a bit confusing for inexperienced R users.]

+ set a folder (any folder) as the working directory using `setwd()`  
+ create a folder called "Data" inside the folder designated as the working directory  
+ download the pertinent datasets from [here](https://www.dropbox.com/sh/rtbs4ji21c9uiy9/AADYpHAWhUxMittAptuq-Apaa?dl=0)
+ place all the files in the downloaded folder in the "Data" folder


## The impact of groundwater pumping on depth to water table {#Demo1}

```{r child = '01_z01_groundwater_pumping.rmd'}
```

## Precision Agriculture {#demo-PA}
```{r child = '01_z02_precision_agriculture.rmd'}
```

## Land Use and Weather {#demo3}

```{r child = '01_z03_landuse_weather.rmd'}
```

## The Impact of Railroad Presence on Corn Planted Acreage {#demo-railroad}

```{r child = '01_z04_railroad_acres.rmd'}
```

## Groundwater use for agricultural irrigation {#demo-gw-ir}

```{r child = '01_z05_groundwater_irrigation.rmd'}
```

## African Economy and Slaves: Nunn 2008 {#demo-slave}

```{r child = '01_z06_nunn_2008.rmd'}
```

## Terrain Ruggedness and Economic Development in Africa: Nunn 2012 {#demo-tri}

```{r child = '01_z07_nunn_2012.rmd'}
```

## TseTse fly suitability index: Alsan 2015 {#demo-tsetse}

```{r child = '01_z08_alsan_2015.rmd'}
```

