### Project Overview

---

**Objective**

+ Understand the impact of railroad on corn planted acreage in Illinois

---

**Datasets**

+ USDA corn planted acreage for Illinois downloaded from the USDA  NationalAgricultural Statistics Service (NASS) QuickStats service using `tidyUSDA` package 
+ US railroads (line data) downloaded from [here](https://catalog.data.gov/dataset/tiger-line-shapefile-2015-nation-u-s-rails-national-shapefile)

---

**Econometric Model**

We will estimate the following model:

$$
  y_i = \beta_0 + \beta_1 RL_i + v_i
$$

where $y_i$ is corn planted acreage in county $i$ in Illinois, $RL_i$ is the total length of railroad, and $v_i$ is the error term.

---

**GIS tasks**

+ Download USDA corn planted acreage by county as a spatial dataset (`sf` object)
  * use `tidyUSDA::getQuickStat()`
+ Import US railroad shape file as a spatial dataset (`sf` object) 
  * use `sf:st_read()`
+ Spatially subset (crop) the railroad data to the geographic boundary of Illinois 
  * use `sf_1[sf_2, ]`
+ Find railroads for each county (cross-county railroad will be chopped into pieces for them to fit within a single county)
  * use `sf::st_intersection()`        
+ Calculate the travel distance of each railroad piece
  * use `sf::st_length()`
* create maps using the `ggplot2` package
  - use `ggplot2::geom_sf()`

---

**Preparation for replication**

+ Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r demo4_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  tidyUSDA, # access USDA NASS data
  sf, # vector data operations
  dplyr, # data wrangling
  ggplot2, # for map creation
  modelsummary, # regression table generation
  keyring # API management
)
```

+ Run the following code to define the theme for map:

```{r define_theme_2, eval = F}
theme_for_map <-
  theme(
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.line = element_blank(),
    panel.border = element_blank(),
    panel.grid.major = element_line(color = "transparent"),
    panel.grid.minor = element_line(color = "transparent"),
    panel.background = element_blank(),
    plot.background = element_rect(fill = "transparent", color = "transparent")
  )
```

### Project Demonstration

We first download corn planted acreage data for 2018 from USDA NASS QuickStat service using `tidyUSDA` package^[In order to actually download the data, you need to obtain the API key [here](https://quickstats.nass.usda.gov/api). Once the API key was obtained, I stored it using `set_key()` from the `keyring` package, which was named "usda_nass_qs_api". In the code to the left, I retrieve the API key using `key_get("usda_nass_qs_api")` in the code. For your replication, replace `key_get("usda_nass_qs_api")` with your own API key.].

```{r get_quicknass, results = "hide", cache = TRUE}
(
  IL_corn_planted <-
    getQuickstat(
      #--- use your own API key here fore replication ---#
      key = key_get("usda_nass_qs_api"),
      program = "SURVEY",
      data_item = "CORN - ACRES PLANTED",
      geographic_level = "COUNTY",
      state = "ILLINOIS",
      year = "2018",
      geometry = TRUE
    ) %>%
    #--- keep only some of the variables ---#
    dplyr::select(year, NAME, county_code, short_desc, Value)
)
```

```{r IL_corn_planted, echo = F}
IL_corn_planted
```

A nice thing about this function is that the data is downloaded as an `sf` object with county geometry with `geometry = TRUE`. So, you can immediately plot it (Figure \@ref(fig:map-il-corn-acreage)) and use it for later spatial interactions without having to merge the downloaded data to an independent county boundary data.

```{r map-il-corn-acreage, fig.cap = "Map of Con Planted Acreage in Illinois in 2018"}
ggplot(IL_corn_planted) +
  geom_sf(aes(fill = Value / 1000)) +
  scale_fill_distiller(name = "Planted Acreage (1000 acres)", palette = "YlOrRd", trans = "reverse") +
  theme(legend.position = "bottom") +
  theme_for_map
```

---

Let's import the U.S. railroad data and reproject to the CRS of `IL_corn_planted`:

```{r Demo5_rail, dependson = "get_quicknass", cache = FALSE}
rail_roads <-
  st_read("Data/tl_2015_us_rails.shp") %>%
  #--- reproject to the CRS of IL_corn_planted ---#
  st_transform(st_crs(IL_corn_planted))
```

Here is what it looks like:

```{r Demo5-rail-plot, fig.cap = "Map of Railroads", cache = TRUE}
ggplot(rail_roads) +
  geom_sf() +
  theme_for_map
```

We now crop it to the Illinois state border (Figure \@ref(fig:Demo5-rail-IL-plot)):

```{r crop_to_IL_run, echo = FALSE, dependson = "get_quicknass", cache = TRUE}
rail_roads_IL <- rail_roads[IL_corn_planted, ]
```

```{r crop_to_IL, eval = FALSE, dependson = "get_quicknass"}
rail_roads_IL <- rail_roads[IL_corn_planted, ]
```

```{r Demo5-rail-IL-plot, fig.cap = "Map of railroads in Illinois", cache = TRUE}
ggplot() +
  geom_sf(data = rail_roads_IL) +
  theme_for_map
```

Let's now find railroads for each county, where cross-county railroads will be chopped into pieces so each piece fits completely within a single county, using `st_intersection()`.

```{r intersect_rails, dependson = "get_quicknass", cache = TRUE}
rails_IL_segmented <- st_intersection(rail_roads_IL, IL_corn_planted)
```

Here are the railroads for Richland County:

```{r map_seg_rail, dependson = "intersect_rails", fig.height = 6}
ggplot() +
  geom_sf(data = dplyr::filter(IL_corn_planted, NAME == "Richland")) +
  geom_sf(
    data = dplyr::filter(rails_IL_segmented, NAME == "Richland"),
    aes(color = LINEARID)
  ) +
  theme(legend.position = "bottom") +
  theme_for_map
```

We now calculate the travel distance (Great-circle distance) of each railroad piece using `st_length()` and then sum them up by county to find total railroad length by county.

```{r rail_total_county}
(
  rail_length_county <-
    mutate(
      rails_IL_segmented,
      length_in_m = as.numeric(st_length(rails_IL_segmented))
    ) %>%
    #--- geometry no longer needed ---#
    st_drop_geometry() %>%
    #--- group by county ID ---#
    group_by(county_code) %>%
    #--- sum rail length by county ---#
    summarize(length_in_m = sum(length_in_m))
)
```

---

We merge the railroad length data to the corn planted acreage data and estimate the model.

```{r merge_data}
reg_data <- left_join(IL_corn_planted, rail_length_county, by = "county_code")
```

```{r rail-reg-table}
lm(Value ~ length_in_m, data = reg_data) %>%
  modelsummary(
    stars = TRUE,
    gof_omit = "IC|Log|Adj|Within|Pseudo"
  )
```

---

