### Project Overview

---

**Objective:**

Create some of the variables used in @nunn2008long (here)[https://scholar.harvard.edu/nunn/publications/long-term-effects-africas-slave-trades]

* the distance variable used as an instrument (distance to the nearest trade center for each country in Africa)   
* the number of slaves for each of the countries in Africa

The tutorial is originally from http://mkudamatsu.github.io/gis_lecture4.html.

---

**Datasets**

* Coast lines of the world
* Country boundary in Africa
* Boundary of ethnic regions in Africa

---

**GIS tasks**

* read an ESRI shape file as an `sf` (spatial) object 
  - use `sf::st_read()`
* simply a spatial object (reduce the number of points representing it)
  - use `rmapshaper::ms_simplify()`
* find the closest point on the boundary of polygons
  - use `sf::st_nearest_points()`
* find the centroid of a polygon
  - use `sf::st_centroid()`
* combine multiple lines into a single line
  - use `sf::st_union()`
* identify the last point of a line
  - use `lwgeom::st_endpoint()`
* calculate the distance between two spatial objects
  - use `sf::st_distance()`
* implement area-weighted spatial interpolation
  - use `sf::st_interpolate_aw()`
* drop geometry from an `sf` object
  - use `sf::st_drop_geometry()`
* convert a regular `data.frame` (non-spatial) with geographic coordinates into an `sf` (spatial) objects
  - use `sf::st_as_sf()` and `sf::st_set_crs()`
* reproject an `sf` object to another CRS
  - use `sf::st_transform()`
  - use `sf::st_join()`
+ create maps 
  * use the `ggplot2` package 

---

**Preparation for replication**

Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r demo1_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  sf, # vector data operations
  tidyverse, # data wrangling
  stars,
  skimr
)
```

### Project Demonstration

# Nunn (2012)

Nunn (2012): https://scholar.harvard.edu/nunn/publications/ruggedness-blessing-bad-geography-africa


```{r Load digital elevation map}
# From https://www.eea.europa.eu/data-and-maps/data/world-digital-elevation-model-etopo5, spatial resolution 5-minute gride, about 10km grid

dem <- stars::read_stars(here::here("data/nunn_2012/GDEM-10km-BW.tif"))

# View stars object
print(dem)

# rename alwdgg.tif to elevation
dem$elevation <- dem$`GDEM-10km-BW.tif`
dem$`GDEM-10km-BW.tif` <- NULL

# Quick plot
plot(dem)
```

Stars raster objects keeps track of the x,y (and sometimes time) position in "bands" and the values in "attributes". The bands keep track of the location of each square and the "attribute" keeps track of the value at that square. Therefore, when we work with stars objects, we only want to work with the attributes. 

The first thing to notice is that the ocean floor is recorded as 0, so we are going to replace an elevation of 0 with `NA`s.

```{r}
dem$elev <- ifelse(dem$elevation == 0, NA, dem$elevation)

plot(dem["elev"])
```


# Nunn (2012) Data Replication

We are going to loop through the raster and calculate the Terrain Ruggedness Index (Riley et. al. (1999)):

$$
	Ruggedness_{r,c} = \sqrt{\sum_{i=r-1}^{r+1} \sum_{j= r-1}^{r+1} (e_{i,j} - e_{r,c})^2}
$$

`stars` is not as developed of a package as sf and therefore ocasionally you will need to use functions/packages designed for the `raster` package. One common one is the `focal` or "moving window" function. It allows you to apply a function to every cell of a raster and as an input bring a matrix of the surrounding values. For example `focal(raster, w= matrix(1, nrow=3, ncol=3), fun= any_function)` will pass to your `any_function` a 3x3 matrix of the raster values centered at the point multiplied by 1 for each cell. You can pass any matrix with values of the weight. For example, `focal(raster, w= matrix(1/9, nrow= 3, ncol=3), fun= sum)` will return the average value of the 3x3 matrix.

The focal function acts weird at the north/south boundaries of the raster. `pad= TRUE, padValue= NA` for example will insert the cells above the northern most cells with NA values when passing to the function. Note focal doesn't do this for the left

```{r Calculate TRI}
library("raster")
demRaster <- as(dem["elev"], "Raster")

calc_ruggedness <- function(matr) {
  # matr is a length 9 matrix
  center <- matr[5]
  sum_squares <- sum((matr - center)^2, na.rm = TRUE)
  return(sqrt(sum_squares))
}

triRaster <- focal(demRaster, w = matrix(1, nrow = 3, ncol = 3), fun = calc_ruggedness, pad = TRUE, padValue = NA)
```

Last we can convert the tri back into a `stars` object and plot it:

```{r}
tri <- st_as_stars(triRaster)

plot(tri)
```





