# Spatial Interactions of Raster Data {#int-RR}

```{r chap6_setup, echo = FALSE, results = "hide"}
library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  # cache = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  tidy = FALSE,
  cache.lazy = FALSE
)

suppressMessages(library(here))
opts_knit$set(root.dir = here())
```

```{r , eval = FALSE, echo = FALSE}
setwd(here())
```

```{r, include = FALSE, cache = FALSE}
#--- load packages ---#
suppressMessages(library(data.table))
suppressMessages(library(exactextractr))
suppressMessages(library(prism))
suppressMessages(library(sf))
suppressMessages(library(terra))
suppressMessages(library(raster))
suppressMessages(library(tidyverse))
suppressMessages(library(DT))
suppressMessages(library(tictoc))
suppressMessages(library(tmap))
suppressMessages(library(parallel))
suppressMessages(library(maps))
```


```{r figure_setup, echo = FALSE}
theme_update(
  axis.title.x = element_text(size = 12, angle = 0, hjust = .5, vjust = -0.3, face = "plain", family = "Times"),
  axis.title.y = element_text(size = 12, angle = 90, hjust = .5, vjust = .9, face = "plain", family = "Times"),
  axis.text.x = element_text(size = 10, angle = 0, hjust = .5, vjust = 1.5, face = "plain", family = "Times"),
  axis.text.y = element_text(size = 10, angle = 0, hjust = 1, vjust = 0, face = "plain", family = "Times"),
  axis.ticks = element_line(size = 0.3, linetype = "solid"),
  # axis.ticks = element_blank(),
  axis.ticks.length = unit(.15, "cm"),
  # axis.ticks.margin = unit(.1,'cm'),
  # axis.text = element_text(margin=unit(.1,'cm')),

  #--- legend ---#
  legend.text = element_text(size = 10, angle = 0, hjust = 0, vjust = 0, face = "plain", family = "Times"),
  legend.title = element_text(size = 10, angle = 0, hjust = 0, vjust = 0, face = "plain", family = "Times"),
  legend.key.size = unit(0.5, "cm"),

  #--- strip (for faceting) ---#
  strip.text = element_text(size = 10, family = "Times"),

  #--- plot title ---#
  plot.title = element_text(family = "Times", face = "bold", size = 12),

  #--- margin ---#
  # plot.margin = margin(0, 0, 0, 0, "cm"),

  #--- panel ---#
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  panel.border = element_rect(fill = NA)
)
```

## Before you start {-}

### Direction for replication {-}

**Datasets**

All the datasets that you need to import are available [here](https://www.dropbox.com/sh/l84zfidaxmjrti9/AAB1GrDRoIlidJ3_zArMN24ua?dl=0). In this chapter, the path to files is set relative to my own working directory (which is hidden). To run the codes without having to mess with paths to the files, follow these steps:

+ set a folder (any folder) as the working directory using `setwd()`  
+ create a folder called "Data" inside the folder designated as the working directory (if you have created a "Data" folder previously, skip this step)
+ download the pertinent datasets from [here](https://www.dropbox.com/sh/l84zfidaxmjrti9/AAB1GrDRoIlidJ3_zArMN24ua?dl=0) 
+ place all the files in the downloaded folder in the "Data" folder

**Packages**

Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r Chap5_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  terra, # handle raster data
  raster, # handle raster data
  exactextractr, # fast extractions
  sf, # vector data operations
  dplyr, # data wrangling
  tidyr, # data wrangling
  data.table, # data wrangling
  prism, # download PRISM data
  tictoc # timing codes
)
```

## Resample

http://132.72.155.230:3838/r/geometric-operations-with-rasters.html

## Mosaic

