# Download and process spatial datasets from within R {#download-data}

## Before you start {-}

There are many publicly available spatial datasets that can be downloaded using R. Programming data downloading using R instead of manually downloading data from websites can save lots of time and also enhances the reproducibility of your analysis. In this section, we will introduce some of such datasets and show how to download and process those data.  

### Direction for replication {-}

**Datasets**

No datasets to download for this Chapter.

**Packages**

+ Run the following code to install or load (if already installed) the `pacman` package, and then install or load (if already installed) the listed package inside the `pacman::p_load()` function.

```{r Chap8_packages}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  stars, # spatiotemporal data handling
  terra, # raster data handling
  raster, # raster data handling
  sf, # vector data handling
  dplyr, # data wrangling
  stringr, # string manipulation
  lubridate, # dates handling
  data.table, # data wrangling
  tidyr, # reshape
  tidyUSDA, # download USDA NASS data
  keyring, # API key management
  FedData, # download Daymet data
  daymetr, # download Daymet data
  ggplot2, # make maps
  tmap, # make maps
  future.apply, # parallel processing
  CropScapeR, # download CDL data
  prism # download PRISM data
)  
```

+ Run the following code to define the theme for map:

```{r define_theme_Chap8, cache = F}
theme_set(theme_bw())

theme_for_map <- theme(
  axis.ticks = element_blank(),
  axis.text= element_blank(), 
  axis.line = element_blank(),
  panel.border = element_blank(),
  panel.grid.major = element_line(color='transparent'),
  panel.grid.minor = element_line(color='transparent'),
  panel.background = element_blank(),
  plot.background = element_rect(fill = "transparent",color='transparent')
)  
```

## USDA NASS QuickStat with `tidyUSDA` {#nass-quick}

There are several packages that lets you download data from the USDA NASS QuickStat. Here we use the `tidyUSDA` package [@tidyUSDA]. A nice thing about [tidyUSDA](https://github.com/bradlindblad/tidyUSDA) is that it gives you an option to download data as an `sf` object, which means you can immediately visualize the data or spatially interact it with other spatial objects.

First thing you want to do is to get an API key from this [website](https://quickstats.nass.usda.gov/api), which you need to actually download data. 

You can download data using `getQuickstat()`. There are number of options you can use to narrow down the scope of the data you are downloading including `data_item`, `geographic_level`, `year`,  `commodity`, and so on. See its [manual](https://cran.r-project.org/web/packages/tidyUSDA/tidyUSDA.pdf) for the full list of parameters you can set. As an example, the code below download corn-related data by county in Illinois for year 2016 as an `sf` object.

```{r il_corn_download, results = 'hide'}
(
IL_corn_yield <- getQuickstat(
    #--- put your API key in place of key_get() ---#
    key = key_get("usda_nass_qs_api"),
    program = "SURVEY",
    commodity = "CORN",
    geographic_level = "COUNTY",
    state = "ILLINOIS",
    year = "2016",
    geometry = TRUE
  )  %>% 
  #--- keep only some of the variables ---#
  dplyr::select(year, NAME, county_code, short_desc, Value)
)
```

```{r il_corn_disp, echo = F}
IL_corn_yield
```

As you can see, it is an `sf` object with `geometry` column due to `geometry = TRUE` option. This means that you can immediately create a map with the data (Figure \@ref(fig:corn-yield-IL)):

```{r corn-yield-IL, fig.cap = "Corn Yield (bu/acre) in Illinois in 2016"}
ggplot() +
  geom_sf(
    data = filter(IL_corn_yield, short_desc == "CORN, GRAIN - YIELD, MEASURED IN BU / ACRE"), 
    aes(fill = Value)
  ) +
  theme_for_map
```

You can download data for multiple states and years at the same time like below (of course, if you want the whole U.S., don't specify the state parameter.). 

```{r il_co_necorn_download}
(
IL_CO_NE_corn <- getQuickstat(
    key = key_get("usda_nass_qs_api"),
    program = "SURVEY",
    commodity = "CORN",
    geographic_level = "COUNTY",
    state = c("ILLINOIS", "COLORADO", "NEBRASKA"),
    year = paste(2014:2018),
    geometry = TRUE
  ) %>% 
  #--- keep only some of the variables ---#
  dplyr::select(year, NAME, county_code, short_desc, Value)
)
```

### Look for parameter values

This package has a function that lets you see all the possible parameter values you can use for many of these parameters. For example, suppose you know you would like irrigated corn yields in Colorado, but you are not sure what parameter value (string) you should supply to the `data_item` parameter. Then, you can do this:^[Of course, you can alternatively go to the QuickStat website and look for the text values.]

```{r see_values_dataitem}
#--- get all the possible values for data_item ---#
all_items <- tidyUSDA::allDataItem 

#--- take a look at the first six ---#
head(all_items)
```

You can use key words like "CORN", "YIELD", "IRRIGATED" to narrow the entire list using `grep()`^[`grep()` is part of the base package, and you do not need to install any package to use it.]: 

```{r }
all_items %>% 
  grep(pattern = "CORN", ., value = TRUE) %>% 
  grep(pattern = "YIELD", ., value = TRUE) %>% 
  grep(pattern = "IRRIGATED", ., value = TRUE)  
```

Looking at the list, we know the exact text value we want, which is the first entry of the vector.

```{r co_corn_download}
(
CO_ir_corn_yield <- getQuickstat(
    key = key_get("usda_nass_qs_api"),
    program = "SURVEY",
    data_item = "CORN, GRAIN, IRRIGATED - YIELD, MEASURED IN BU / ACRE",
    geographic_level = "COUNTY",
    state = "COLORADO",
    year = "2018",
    geometry = TRUE
  ) %>% 
  #--- keep only some of the variables ---#
  dplyr::select(year, NAME, county_code, short_desc, Value)
)
```

Here is the complete list of functions that gives you the possible values of the parameters for `getQuickstat()`. 

```{r eval = F}
tidyUSDA::allCategory 
tidyUSDA::allSector 
tidyUSDA::allGroup 
tidyUSDA::allCommodity 
tidyUSDA::allDomain 
tidyUSDA::allCounty 
tidyUSDA::allProgram 
tidyUSDA::allDataItem 
tidyUSDA::allState 
tidyUSDA::allGeogLevel 
```

### Caveats

You cannot retrieve more than $50,000$ (the limit is set by QuickStat) rows of data. The query below requests much more than $50,000$ observations, and fail. In this case, you need to narrow the search and chop the task into smaller tasks. 

```{r cav_1, error = TRUE}
many_states_corn <- getQuickstat(
    key = key_get("usda_nass_qs_api"),
    program = "SURVEY",
    commodity = "CORN",
    geographic_level = "COUNTY",
    state = c("ILLINOIS", "COLORADO", "NEBRASKA", "IOWA", "KANSAS"),
    year = paste(1995:2018),
    geometry = TRUE
  ) 
```

Another caveat is that the query returns an error when there is no observation that satisfy your query criteria. For example, even though "CORN, GRAIN, IRRIGATED - YIELD, MEASURED IN BU / ACRE" is one of the values you can use for `data_item`, there is no entry for the statistic in Illinois in 2018. Therefore, the following query fails.

```{r cav_2, error = TRUE}
many_states_corn <- getQuickstat(
    key = key_get("usda_nass_qs_api"),
    program = "SURVEY",
    data_item = "CORN, GRAIN, IRRIGATED - YIELD, MEASURED IN BU / ACRE",
    geographic_level = "COUNTY",
    state = "ILLINOIS",
    year = "2018",
    geometry = TRUE
  ) 
```

<!-- 
#/*=================================================*/
#' # CDL
#/*=================================================*/
-->

## CDL with `CropScapeR` {#CropScapeR}

The Cropland Data Layer (CDL) is a data product produced by the National Agricultural Statistics Service of U.S. Department of Agriculture. CDL provides geo-referenced, high accuracy, 30 (after 2007) or 56 (in 2006 and 2007) meter resolution, crop-specific cropland land cover information for up to 48 contiguous states in the U.S. from 1997 to the present. This data product has been extensively used in agricultural research. CropScape is an interactive Web CDL exploring system (<https://nassgeodata.gmu.edu/CropScape/>), and it was developed to query, visualize, disseminate, and analyze CDL data geospatially through standard geospatial web services in a publicly accessible on-line environment (Han et al., 2012). 

This section shows how to use the `CropScapeR` package [@cropscaper] to download and explore the CDL data. The package implements some of the most useful geospatial processing services provided by the CropScape, and it allows users to efficiently process the CDL data within the `R` environment. Specifically, the `CropScapeR` package provides four functions that implement different kinds of geospatial processing services provided by the CropScape. This section introduces these functions while providing some examples. `GetCDLData()` in particular is the most important function as it lets you download the raw CDL data. The other functions provide the users with the CDL data summarized or transformed in particular manners that may suit the need of some users.

**Note**: There is a known problem with Mac users requesting CDL data services using the CropScape API, which causes errors when using the functions provided by the package. Please see section \@ref(mac-problem) for a workaround.

---

The `CropScapeR` package can be installed directly from 'CRAN':    
```{r, eval  = F}
install.packages("CropScapeR")
```

The development version of the package can be downloaded from [its GitHub website](https://github.com/cbw1243/CropScapeR) using the following codes:   

```{r, eval  = F}
library(devtools)
devtools::install_github("cbw1243/CropScapeR")  
```

Let's load the package.

```{r CropScapeR-load, eval = F}
library(CropScapeR)  
```


```{r, include=F, message=F}
httr::set_config(httr::config(ssl_verifypeer = 0L))

data <- GetCDLData(aoi = 17019, year = 2018, type = 'f', save_path = 'data.tif')
cdl_Champaign <- GetCDLData(aoi = 17019, year = 2018, type = 'f', save_path = 'ch.tif')

cdl_sf <- GetCDLData(aoi = 17019, year = 2018, type = 'f', format = "sf", save_path = 'cdl_sf.tif')
```

---

Acknowledgment: The development of the `CropScapeR` package was supported by USDA-NRCS Agreement No. NR193A750016C001 through the Cooperative Ecosystem Studies Units network. Any opinions, findings, conclusions, or recommendations expressed are those of the author(s) and do not necessarily reflect the view of the U.S. Department of Agriculture.

### `GetCDLData`: Download the CDL data as raster data

<!-- The `GetCDLData` function fetches the cropland land cover data directly from the CDL. This is done in two steps: use the GET function from the httr package to send data requests; then use the `raster` function from the `raster` package to read the requested CDL data (a raster TIF file). 
 -->

`GetCDLData()` allows us to obtain CDL data for any Area of Interest (AOI) in a given year. It requires three parameters to make a valid data request: 

+ `aoi`: Area of Interest (AOI).     
+ `year`: Year of the data to request.    
+ `type`: Type of AOI.    

The following AOI-type combinations are accepted:

+ any spatial object as an `sf` or `sfc` object - `type = "b"`
+ county (defined by a 5-digit county FIPS code) - `type = "f"`
+ state (defined by a 2-digit state FIPS code) - `type = "f"`
+ bounding box (defined by four corner points) - `type = "b"`
+ polygon area (defined by at least three coordinates) - `type = "ps"`
+ single point (defined by a coordinate) - `type = "p"`

This section discusses how to download data for an `sf` object, county, and state as they are likely to be the most common AOI. See the package github site (<https://github.com/cbw1243/CropScapeR>) to see how the other options work. 

#### Downloading CDL data for `sf`, county, and state

**Downloading CDL data for `sf`**

Let's download the 2018 CDL data for the area that covers Champaign, Vermilion, Ford, and Iroquois Counties in Illinois (a map below).

```{r IL_4_dips, eval = F}
library(tigris)

#--- get the sf for the four counties  ---# 
IL_4_county <- counties(state = "Illinois", cb = TRUE) %>% 
  st_as_sf() %>% 
  filter(NAME %in% c("Champaign", "Vermilion", "Ford", "Iroquois")) 
```

```{r IL_4_dips_run, include = F}
library(tigris)
 
IL_county <- counties(state = "Illinois", cb = TRUE) %>% 
  st_as_sf() 

(
IL_county_4 <- filter(IL_county, NAME %in% c("Champaign", "Vermilion", "Ford", "Iroquois"))
)
```

```{r IL4-map}
ggplot() +
  geom_sf(data = IL_county) +
  geom_sf(data = IL_county_4, fill = "lightblue") +
  theme_void()
```

When you use an `sf` object for `aoi`, CDL data will be downloaded for the bounding box (this is why `type = "b"`) that encompasses the entire geographic area of the `sf` object irrespective of the type of objects in the `sf` object (whether they are points, polygons, lines). In this case, CDL data is downloaded for the red area in the map below. 

```{r IL4-aoi-map}
ggplot() +
  geom_sf(data = IL_county) +
  geom_sf(data = st_as_sfc(st_bbox(IL_county_4)), fill = "red", alpha = 0.4) +
  theme_void() 
```

Let's download CDL data for the four counties: 

```{r IL4-download-disp, eval = F}
(
cdl_IL_4 <- GetCDLData(
  aoi = IL_county_4, 
  year = "2018", 
  type = "b" 
  )
)

``` 

```{r IL4-read, echo = F}
# cdl_IL_4 <- GetCDLData(
#   aoi = IL_county_4, 
#   year = "2018", 
#   type = "b",
#   save_path = "IL4.tif"
# )
(
cdl_IL_4 <- raster("./Data/IL4.tif")
)

``` 

As you can see, the downloaded data is a `RasterLayer` object^[an object class defined by the `raster` package. See \@ref(raster-basics).]. Note that the CDL data uses the Albers equal-area conic projection. 

```{r proj-IL4}
projection(cdl_IL_4) 
```

Take a look at the downloaded CDL data.

```{r plot-IL4}
plot(cdl_IL_4)
```

If you do not want to have values outside of the `sf` object, you can use `raster::mask()` to turn them into NA as follows:

```{r mask-IL4}
cdl_IL_4_masked <- IL_county_4 %>% 
  #--- change the CRS first to that of the raster data ---#
  st_transform(., projection(cdl_IL_4)) %>% 
  #--- mask the values outside the sf (turn them into NA) ---#
  raster::mask(cdl_IL_4, .)  
```

As you can see below, values outside the four counties are now `NA` (black area):

```{r plot-IL4-mask}
plot(cdl_IL_4_masked)
```


---

**Downloading CDL data for county** 

The following code makes a request to download the CDL data for Champaign county in Illinois in 2018. 

```{r eval = F}
(
cdl_Champaign <- GetCDLData(aoi = 17019, year = 2018, type = 'f')
)
```

```{r echo = F}
(
cdl_Champaign <- raster("ch.tif")
)

```

In the above code, the FIPS code for Champaign County (17019) was supplied to the `aoi` option. Because a county is used here, the `type` argument is specified as 'f'. 

```{r fig.width=3, fig.height=3, fig.align='center'}
plot(cdl_Champaign) 
```

---

**Downloading CDL data for state** 

The following code makes a request to download the CDL data for the state of Illinois in 2018. 

```{r eval = F}
(
cdl_IL <- GetCDLData(aoi = 17, year = 2018, type = 'f')
)
```

```{r include = F}
# (
# cdl_IL <- GetCDLData(aoi = 17, year = 2018, type = 'f', save_path = "IL_state.tif", tol_time = 1000)
# )

(
cdl_IL <- raster("./Data/il.tif")
)

```

In the above code, the state FIPS code for Illinois ($17$) was supplied to the `aoi` option. Because a county is used here, the `type` argument is specified as 'f'. 

```{r fig.width=3, fig.height=3, fig.align='center'}
plot(cdl_IL) 
```

#### Other format options

---

**GeoTiff**

You could save the downloaded CDL data as a tif file by adding `save_path =` option to `GetCDLData()` as follows:

```{r eval = F}
(
cdl_IL_4 <- GetCDLData(
  aoi = IL_county_4, 
  year = "2018", 
  type = "b",
  save_path = "/Data/IL_4.tif"
  )
)

``` 

With this code, the downloaded data will be saved as "IL_4.tif" in the "Data" folder residing in the current working directory. 

---

**sf**

The `GetCDLData` function lets you download CDL data as an `sf` of points, where the coordinates of the points are the coordinates of the centroid of the raster cells. This can be done by adding `format = sf` as an option. 

```{r eval = F}
(
cdl_sf <- GetCDLData(aoi = 17019, year = 2018, type = 'f', format = "sf")
)
``` 

```{r include = F}
cdl_sf
```

The first column (`value`) is the crop code. Of course, you can manually convert a RasterLayer to an `sf` of points as follows:

```{r raster-sf-manu}
as.data.frame(cdl_Champaign, xy = TRUE) %>% 
  #--- to sf consisting of points ---#
  st_as_sf(coords = c("x", "y")) %>% 
  #--- Albert conic projection ---#
  st_set_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +units=m +no_defs")  
```

The `format = sf` option makes `GetCDLData()` do this conversion internally for those who want CDL data as an `sf` consisting of points instead of a `RasterLayer`.

### Data processing after downloading data

The downloaded raster data itself is not readily usable immediately for your economic analysis. Typically the variable of interest is the frequency of land use types or their shares. You can use `raster::freq()` to get the frequency (the number of raster cells) of each land use type.

```{r get-freq}
(
crop_freq <- freq(cdl_Champaign)
)
```

Clearly, once frequencies are found, you can easily get shares as well:

```{r get-share}
(
crop_data <-crop_freq %>% 
  #--- matrix to data.frame ---#
  data.frame(.) %>% 
  #--- find share ---#
  mutate(share = count/sum(count))
)

```

At this point, the data does not tell you which value corresponds to which crop. To find the crop names associated with the crop codes (`value`), you can get the reference table using `data(linkdata)` from the `CropScapeR` package.^[Alternatively, you can download the reference table (a EXCEL file) at <https://www.nass.usda.gov/Research_and_Science/Cropland/docs/cdl_codes_names.xlsx>.] 

```{r}
#--- load the crop code reference data ---#
data("linkdata")
```

```{r echo = F}
library(DT)
datatable(linkdata)  
```

You can merge the two data sets using `value` from the CDL data and `MasterCat` from `linkdata` as the merging keys:


```{r }
(
crop_data <- dplyr::left_join(crop_data, linkdata, by = c('value' = 'MasterCat'))
)
``` 

`NoData` in `Crop` corresponds to the black area in the above figure, which is the portion of the raster data that does not overlap with the boundary of Champaign County. These points with `NoData` can be removed by using the `filter` function. 

### Other forms of CDL data

Instead of downloading the raw CDL data, CropScape provides an option to download summarized CDL data.

+ `GetCDLComp`: request data on land use changes
+ `GetCDLStat`: get acreage estimates from the CDL
+ `GetCDLImage`: download the image files of the CDL data

These may come handy if they satisfy your needs because you can skip post-downloading processing steps. 

---

**`GetCDLComp()`**: request data on land use changes

The `GetCDLComp` function allows users to request data on changes in land cover over time from the CDL. Specifically, this function returns acres changed from one crop category to another crop category between two years for a user-defined AOI. 

Let's see an example. The following codes request data on acreage changes in land cover for Champaign County (FIPS = 17019) from 2017 (`year1 = 2017`) to 2018 (`year2 = 2018`). 

```{r getcdl-comp, eval = F}
(
data_change <- GetCDLComp(aoi = '17019', year1 = 2017, year2 = 2018, type = 'f')
)
```

```{r echo = F}
readRDS("cdl_omp.rds")
```

What is returned is a `data.frame` (`data.table`) that has 5 columns. The columns "From" and "To" are crop names. The column "Count" is the pixel count, and "Acreage" is the acres corresponding to the pixel counts. The last column "aoi" is the selected AOI. The first row of the returned data table shows the acreage (i.e., 40,362 acres) of continuous corn during 2017 and 2018. The third row shows the acreage (i.e., 240,506 acres) rotated from corn to soybeans during 2017 and 2018.   

Remember that the spatial resolution changes from 56 meters to 30 meters starting 2008. This means that when data is requested for land use changes from 2007 to 2008, two CDL raster layers have different spatial resolutions. Consequently, CropScape API fails to resolve the issue and return an error message saying "Mismatch size of file 1 and file 2." The `GetCDLComp()` function manually resolves this problem by resampling two CDL raster files using the nearest neighbor resampling technique such that both rasters have the same spatial resolutions. The finer resolution raster is downscaled to the lower resolution. Then, the resampled raster layers are merged together to calculate cropland changes. Users can turn off this default behavior by adding `manual_try = FALSE` option. In this case, an error message from CropScape API will be returned with no land use changes results.

```{r resolu-change, error = TRUE}
data_change <- GetCDLComp(aoi = '17019', year1 = 2007, year2 = 2008, type = 'f', `manual_try` = FALSE) 
```

---

**`GetCDLStat()`**: get acreage estimates from the CDL      

The `GetCDLStat` function allows users to get acreage by land cover category for a user defined AOI in a year. For example, the following codes request data on acreage by land cover categories for Champaign County in Illinois in 2018. You can see that the pixel counts are already converted to acres and category names are attached.   

```{r getcdl-stat, eval = F}
(
data_stat <- GetCDLStat(aoi = 17019, year = 2018, type = 'f')
)
``` 

```{r echo = F}
# saveRDS(data_stat, "cdl_stat.rds") 
readRDS("cdl_stat.rds") 
```

---

**`GetCDLImage()`**: Download the image files of the CDL data
The `GetCDLImage` function allows users to download the image files of the CDL data. This function is very similar to the `GetCDLData` function, except that image files are returned. This function can be helpful if you only want to look at the picture of the CDL data. By default, the picture is saved as the 'png' format. You can also save it in the 'kml' format. 

```{r getcdl-image, eval = F}
GetCDLImage(aoi = 17019, year = 2018, type = 'f', verbose = F)
``` 

### SSL certificate problem on Mac {#mac-problem}

SSL refers to the Secure Sockets Layer, and SSL certificate displays important information for verifying the owner of a website and encrypting web traffic with SSL/TLS for securing connection. It is a known problem that Mac users encounter the following error when `GetCDLData()` is used: 'SSL certificate problem: SSL certificate expired'. As the name suggests, this is because the CropScape server has an expired certificate. While this affects the Mac users, Windows users should not expect this issue. 

To avoid the problem for Mac users, the `CropScapeR` has a workaround that involves downloading the GeoTiff file for the requested AOI first, and then read the file using the `raster()` function as a `RasterLayer`.^[Specifically, at step 1, specify in `R` that you want to skip the certificate validation when making the httr GET request. At step 2, download the raster TIF data into your local drive using the `download.file` function with `wget`, and then read the downloaded raster file using the `raster` function.] 

You first need to run the following code before requesting data from CDL.

```{r ssl}
#--- Skip the SSL check ---#
httr::set_config(httr::config(ssl_verifypeer = 0L)) 
```

Now, you can download CDL data by specifying the file path with the `save_path` option like below:

```{r eval = F}

#--- Download the raster TIF file into specified path, also read into R ---#
data <- GetCDLData(aoi = 17019, year = 2018, type = 'f', save_path = "ch.tif")
```

Hopefully, this problem is fixed by the maintainer of CropScape so that this workaround is not necessary for Mac users.

<!-- ### An example: parallelized processing of CDL data -->

<!-- 
#/*=================================================*/
#' # PRISM
#/*=================================================*/
-->

## PRISM with `prism` {#download-prism}

### Basics

[PRISM dataset](https://prism.oregonstate.edu/) provide model-based estimates of precipitation, tmax, and tmin for the U.S. at the 4km by 4km spatial resolution. Here, we use `get_prism_dailys()` from the `prism` package [@prism] to download daily data. Here is its general syntax:

```{r prism_syntax, eval = F}
#--- NOT RUN ---#
get_prism_dailys(
  type = variable type,
  minDate = starting date as character,
  maxDate = ending date as character,
  keepZip = TRUE or FALSE
) 
```

The variables types you can select from is "ppt" (precipitation), "tmean" (mean temperature), "tmin" (minimum temperature), and "tmax" (maximum temperature). For `minDate` and `maxDate`, the dates must be specified in a specific format of "YYYY-MM-DD". `keepZip = FALSE` does not keep the zipped folders of the downloaded files as the name suggests. 

Before you download PRISM data using the function, it is recommended that you set the path to folder in which the downloaded PRISM will be stored using `options(prism.path = "path")`. For example, the following set the path to "Data/PRISM/" relative to the current working directory.

```{r option-prism, eval = F}
options(prism.path = "./Data/PRISM/")
```

The following code downloads daily precipitation data from January 1, 2000 to Jan 10, 2000.

```{r ex, eval = F}
#--- NOT RUN ---#
get_prism_dailys(
  type = "ppt",
  minDate = "2000-01-01",
  maxDate = "2000-01-10",
  keepZip = FALSE
) 
```

When you download data using the above code, you will notice that it creates one folder for one day. For example, for precipitation data for "2000-01-01", you can get the path to the downloaded file as follows: 

```{r rule-folder}
var_type <- "ppt" # variable type
dates_prism_txt <- str_remove_all("2000-01-01", "-") # date without dashes

#--- folder name ---#
folder_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil") 

#--- file name of the downloaded data inside the above folder ---#
file_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil.bil") 

#--- path to the file relative to the designated data folder (here, it's "./Data/PRISM/") ---#
(
file_path <- paste0("./Data/PRISM/", folder_name, "/", file_name)
)
```

We can then easily read the data using `terra::rast()` or `stars::read_stars()` if you prefer the `stars` way.

```{r read-prism-download}
#--- as SpatRaster ---#
(
prism_2000_01_01_sr <- rast(file_path) 
)

#--- as stars ---#
(
prism_2000_01_01_stars <- read_stars(file_path) 
)
```

Here is a quick visualization of the data (Figure \@ref(fig:quick-viz)):

```{r quick-viz, fig.cap = "PRISM precipitation data for January 1, 2000"}
plot(prism_2000_01_01_stars)
```

As you can see, the data covers the entire contiguous U.S.


### Download daily PRISM data for many years and build your own datasets 

Here, an example of how to create your own sets of PRISM datasets is presented. Creating such datasets and have them locally can be useful if you expect to use the data for many different projects in the future. 

Suppose we are interested in saving daily PRISM precipitation data by year-month from 1980 to 2018. We will write a loop that loops over all the year-month combinations. Before writing a loop, let's work on the code for a particular year-month combination: 1990-December. However, we will write codes in a way that can be easily translated into a looped operations later. Specifically, we will define the following variables and use them as if they are variables to be looped over in a loop.

```{r }
#--- month to work on ---#  
temp_month <- 12

#--- year to work on ---#  
temp_year <- 1990
```

We first need to set the path to the folder in which daily PRISM files will be downloaded. 

```{r set-path}
#--- set your own path ---#
options(prism.path = "./Data/PRISM/") 
```

We then set the start and end dates for `get_prism_dailys()`. 

```{r set-dates}
#--- starting date of the working month-year ---#
(
start_date <- dmy(paste0("1/", temp_month, "/", temp_year))
)

#--- ending date: add a month and then go back 1 day ---#
(
end_date <- start_date %m+% months(1) - 1
)
```

We now download PRISM data for the year-month we are working on.

```{r get-prism-daily-month, eval = F}
#--- download daily PRISM data for the working month-year ---#
get_prism_dailys(
  type = "ppt",
  minDate = as.character(start_date),
  maxDate = as.character(end_date),
  keepZip = FALSE
) 
```
 
Once all the data are downloaded, we will read and import them onto R. To do so, we will need the path to all the downloaded files.

```{r get-file-path}
#--- list of dates of the working month-year ---#
dates_ls <- seq(start_date, end_date, "days") 

#--- remove dashes ---#
dates_prism_txt <- str_remove_all(dates_ls, "-")

#--- folder names ---#
folder_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil") 
#--- the file name of the downloaded data ---#
file_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil.bil") 
#--- complete path to the downloaded files ---#
(
file_path <- paste0("./Data/PRISM/", folder_name, "/", file_name) 
)
```

We now read them as a `stars` object, set the third dimension as date using `Dates` object class, and then save it as an R dataset (This will ensure that date dimensions is kept. See Chapter \@ref(read-write-stars)). 

```{r save-as-stars, eval = F}
(
#--- combine all the PRISM files as stars ---#
temp_stars <- read_stars(file_path, along = 3)
  #--- set the third dimension as data ---# 
  st_set_dimensions("band", values = dates_ls, name = "date")
)

#--- save the stars as an rds file ---#
saveRDS(
  temp_stars, 
  paste0("./Data/PRISM/PRISM_", var_type, "_y", temp_year, "_m", temp_month, ".rds")
)  
```

You could alternatively read the files into a `SpatRaster` object and save it data as a GeoTIFF file.

```{r save-as-tiff, eval = F}
(
#--- combine all the PRISM files as a RasterStack ---#
temp_stars <- terra::rast(file_path) 
)

#--- save as a multi-band GeoTIFF file ---#
writeRaster(temp_stars, paste0("./Data/PRISM/PRISM_", var_type, "_y", temp_year, "_m", temp_month, ".tif"), overwrite = T) 
```

Note that this option of course does not have date as the third dimension. Moreover, the RDS file above takes up only 14 Mb, while the tif file occupies 108 Mb.

Finally, if you would like, you can delete all the individual PRISM files:

```{r remove-prism, eval = F}
#--- delete all the downloaded files ---#
unlink(paste0("./Data/PRISM/", folder_name), recursive = TRUE) 
```

---

Okay, now that we know what to do with a particular year-month combination, we can easily write a loop to go over all the year-month combinations for the period of interest. Since all the processes we observed above for a single year-month combination is embarrassingly parallel, it is easy to parallelize using `future.apply::future_lapply()` or `parallel::mclapply()` (Linux/Mac users only). Here we use `future_lapply()`. Let's first get the number of logical cores.

```{r get_cores}
library(parallel)
num_cores <- detectCores() 

plan(multiprocess, workers = num_cores)
```

The following function goes through all the steps we saw above for a single year-month combination.

```{r def-function}
#--- define a function to download and save PRISM data stacked by month ---#
get_save_prism <- function(i, var_type) {

  print(paste0("working on ", i))

  temp_month <- month_year_data[i, month] # working month
  temp_year <- month_year_data[i, year] # working year
  
  #--- starting date of the working month-year ---#
  start_date <- dmy(paste0("1/", temp_month, "/", temp_year))
  #--- end date ---#
  end_date <- start_date %m+% months(1) - 1

  
  #--- download daily PRISM data for the working month-year ---#
  get_prism_dailys(
    type = var_type,
    minDate = as.character(start_date),
    maxDate = as.character(end_date),
    keepZip = FALSE
  ) 

  #--- list of dates of the working month-year ---#
  dates_ls <- seq(start_date, end_date, "days") 

  #--- remove dashes ---#
  dates_prism_txt <- str_remove_all(dates_ls, "-")

  #--- folder names ---#
  folder_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil") 
  #--- the file name of the downloaded data ---#
  file_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil.bil") 
  #--- complete path to the downloaded files ---#
  file_path <- paste0("./Data/PRISM/", folder_name, "/", file_name)

  #--- combine all the PRISM files as a RasterStack ---#
  temp_stars <- stack(file_path) %>% 
    #--- convert to stars ---#
    st_as_stars() %>% 
    #--- set the third dimension as data ---# 
    st_set_dimensions("band", values = dates_ls, name = "date")

  #--- save the stars as an rds file ---#
  saveRDS(
    temp_stars, 
    paste0("./Data/PRISM/PRISM_", var_type, "_y", temp_year, "_m", temp_month, ".rds")
  )

  #--- delete all the downloaded files ---#
  unlink(paste0("./Data/PRISM/", folder_name), recursive = TRUE)
} 
```

We then create a `data.frame` of all the year-month combinations: 

```{r data_prep_par_ml}
(
#--- create a set of year-month combinations to loop over ---#
month_year_data <- expand.grid(month  = 1:12, year = 1990:2018) %>% 
  data.table()
)
```

We now do parallelized loop over all the year-month combinations (by looping over the rows of the `month_year_data`):

```{r final-loop, eval = F}
#--- run the above code in parallel ---#
future_lapply(
  1:nrow(month_year_data), 
  function (x) get_save_prism(x, "ppt")
)
```

That's it. Of course, you can do the same thing for tmax by this:

```{r final-loop-tmax, eval = F}
#--- run the above code in parallel ---#
future_lapply(
  1:nrow(month_year_data), 
  function (x) get_save_prism(x, "tmax")
)
```

Now that you have PRISM datasets, you can extract values from the raster layers for vector data for your analysis, which is covered extensively in Chapters \@ref(int-RV) and \@ref(stars-basics) (for `stars` objects). 

---

If you want to save the data by year (each file would be about 168 Mb). You could do this.

```{r def-function-y, eval = F}
#--- define a function to download and save PRISM data stacked by year ---#
get_save_prism_y <- function(temp_year, var_type) {

  print(paste0("working on ", temp_year))
  
  #--- starting date of the working month-year ---#
  start_date <- dmy(paste0("1/1/", temp_year))
  #--- end date ---#
  end_date <- dmy(paste0("1/1/", temp_year + 1)) - 1

  #--- download daily PRISM data for the working month-year ---#
  get_prism_dailys(
    type = var_type,
    minDate = as.character(start_date),
    maxDate = as.character(end_date),
    keepZip = FALSE
  ) 

  #--- list of dates of the working month-year ---#
  dates_ls <- seq(start_date, end_date, "days") 

  #--- remove dashes ---#
  dates_prism_txt <- str_remove_all(dates_ls, "-")

  #--- folder names ---#
  folder_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil") 
  #--- the file name of the downloaded data ---#
  file_name <- paste0("PRISM_", var_type, "_stable_4kmD2_", dates_prism_txt, "_bil.bil") 
  #--- complete path to the downloaded files ---#
  file_path <- paste0("./Data/PRISM/", folder_name, "/", file_name)

  #--- combine all the PRISM files as a RasterStack ---#
  temp_stars <- stack(file_path) %>% 
    #--- convert to stars ---#
    st_as_stars() %>% 
    #--- set the third dimension as data ---# 
    st_set_dimensions("band", values = dates_ls, name = "date")

  #--- save the stars as an rds file ---#
  saveRDS(
    temp_stars, 
    paste0("./Data/PRISM/PRISM_", var_type, "_y", temp_year, ".rds")
  )

  #--- delete all the downloaded files ---#
  unlink(paste0("./Data/PRISM/", folder_name), recursive = TRUE)
} 

#--- run the above code in parallel ---#
future_lapply(
  1990:2018, 
  function (x) get_save_prism_y(x, "tmax")
)
```

## Daymet with `daymetr` and `FedData`

For this section, we use the `daymetr` [@daymetr] and `FedData` packages [@feddata].

```{r daymetr_load}
library(daymetr)  
library(FedData)  
```

Daymet data consists of "tiles," each of which consisting of raster cells of 1km by 1km. Here is the map of the tiles (Figure \@ref(fig:daymet-tiles))

```{r daymet-tiles, fig.cap = "Daymet Tiles", echo = F}
library(maps)
US_map <- st_as_sf(map(database = "state", plot = FALSE, fill = TRUE)) %>% 
  st_transform(st_crs(tile_outlines)) 

#--- tiles ---#
# tm_shape(tile_outlines) +
#   tm_polygons(col = "blue", alpha = 0.5) +
# tm_shape(US_map) +
#   tm_polygons(col = "red", alpha = 0.2) 

ggplot() +
  geom_sf(data = st_as_sf(tile_outlines), fill = NA, size = 0.7) +
  geom_sf(data = US_map, fill = "red", alpha = 0.2) +
  theme_for_map
```

Here is the list of weather variables:  

+ vapor pressure 
+ minimum and maximum temperature 
+ snow water equivalent 
+ solar radiation 
+ precipitation  
+ day length 

So, Daymet provides information about more weather variables than PRISM, which helps to find some weather-dependent metrics, such as evapotranspiration. 

The easiest way find Daymet values for your vector data depends on whether it is points or polygons data. For points data, `daymetr::download_daymet()` is the easiest because it will directly return weather values to the point of interest. Internally, it finds the cell in which the point is located, and then return the values of the cell for the specified length of period. `daymetr::download_daymet()` does all this for you. For polygons, we need to first download pertinent Daymet data for the region of interest first and then extract values for each of the polygons ourselves, which we learned in Chapter \@ref(int-RV). Unfortunately, the netCDF data downloaded from `daymetr::download_daymet_ncss()` cannot be easily read by the `raster` package nor the `stars` package. On the contrary, `FedData::get_daymet()` download the the requested Daymet data and save it as a `RasterBrick` object, which can be easily turned into `stars` object using `st_as_stars()`.

### For points data 

For points data, the easiest way to associate daily weather values to them is to use `download_daymet()`. `download_daymet()` can download daily weather data for a single point at a time by finding which cell of a tile the point is located within. 

Here are key parameters for the function:

+ lat: latitude
+ lon: longitude
+ start: start_year
+ end: end_year
+ internal: `TRUE` (dafault) or `FALSE`

For example, the code below downloads daily weather data for a point (lat = $36$, longitude = $-100$) starting from 2000 to 2002 and assigns the downloaded data to `temp_daymet`.  

```{r download_daymet_point}
#--- download daymet data ---#
temp_daymet <- download_daymet(
  lat = 36,
  lon = -100,
  start = 2000,
  end = 2002
) 

#--- structure ---#
str(temp_daymet)
```

As you can see, `temp_daymet` has bunch of site information other than the download Daymet data. 

```{r get_data_part}
#--- get the data part ---#
temp_daymet_data <- temp_daymet$data

#--- take a look ---#
head(temp_daymet_data)
```

As you might have noticed, `yday` is not the date of each observation, but the day of the year. You can easily convert it into dates like this:

```{r }
temp_daymet_data <- mutate(temp_daymet_data, date = as.Date(paste(year, yday, sep = "-"), "%Y-%j"))
```

One dates are obtained, you can use the `lubridate` package to extract day, month, and year using `day()`, `month()`, and `year()`, respectively.

```{r dates}
library(lubridate)

temp_daymet_data <- mutate(temp_daymet_data,
  day = day(date),
  month = month(date),
  #--- this is already there though ---#
  year = year(date)
)

#--- take a look ---#
dplyr::select(temp_daymet_data, year, month, day) %>%  head
```

This helps you find group statistics like monthly precipitation. 

```{r monthly_prcp}
temp_daymet_data %>% 
  group_by(month) %>% 
  summarize(prcp = mean(prcp..mm.day.))
```

Downloading Daymet data for many points is just applying the same operations above to them using a loop. Let's create random points within California and get their coordinates. 

```{r echo = F}
set.seed(389548) 
```

```{r get_random_poitns}
random_points <- maps::map("state", "california", plot = FALSE, fill = TRUE) %>% 
  st_as_sf() %>% 
  #--- 10 points ---#
  st_sample(10) %>% 
  #--- get the coordinates ---#
  st_coordinates() %>% 
  #--- as tibble (data.frame) ---#
  as_tibble() %>% 
  #--- assign site id ---#
  mutate(site_id = 1:n())
```

To loop over the points, you can first write a function like this:

```{r }
get_daymet <- function(i){

  temp_lat <- random_points[i, ] %>% pull(Y)
  temp_lon <- random_points[i, ] %>% pull(X)
  temp_site <- random_points[i, ] %>% pull(site_id)

  temp_daymet <- download_daymet(
    lat = temp_lat,
    lon = temp_lon,
    start = 2000,
    end = 2002
  ) %>% 
  #--- just get the data part ---#
  .$data %>% 
  #--- convert to tibble (not strictly necessary) ---#
  as_tibble() %>% 
  #--- assign site_id so you know which record is for which site_id ---#
  mutate(site_id = temp_site) %>% 
  #--- get date from day of the year ---#
  mutate(date = as.Date(paste(year, yday, sep = "-"), "%Y-%j"))

  return(temp_daymet)
}  
```

Here is what the function returns for the 1st row of `random_points`:

```{r first_daymet}
get_daymet(1)
```

You can now simply loop over the rows.


```{r loop_daymet, eval = F}
(
daymet_all_points <- lapply(1:nrow(random_points), get_daymet) %>% 
  #--- need to combine the list of data.frames into a single data.frame ---#
  bind_rows()
)
```

```{r disp_daymet, echo = F}
# saveRDS(daymet_all_points, "./Data/daymet_all_points.rds")
(
daymet_all_points <- readRDS("./Data/daymet_all_points.rds")
)
```

Or better yet, you can easily parallelize this process as follows (see Chapter \@ref(par-comp) if you are not familiar with parallelization in R):

```{r par_daymet, eval = F}
library(future.apply)    
library(parallel)

#--- parallelization planning ---#
plan(multiprocess, workers = detectCores() - 1)

#--- parallelized lapply ---#
daymet_all_points <- future_lapply(1:nrow(random_points), get_daymet) %>% 
  #--- need to combine the list of data.frames into a single data.frame ---#
  bind_rows()
```  

### For polygons data {#daymet-poly}

Suppose you are interested in getting Daymet data for select counties in Michigan (Figure \@ref(fig:michi-fig)).

```{r MI_counties}
#--- entire MI ---#
MI_counties <- maps::map("county", "michigan", plot = FALSE, fill = TRUE) %>% 
  st_as_sf()  

#--- select counties ---#
MI_counties_select <- filter(MI_counties, ID %in% c("michigan,luce", "michigan,chippewa", "michigan,mackinac"))
```

```{r michi-fig, fig.cap = "Select Michigan counties for which we download Daymet data", echo = F}
tm_shape(MI_counties) +
  tm_polygons(fill = NA) +
tm_shape(MI_counties_select) +
  tm_polygons(col = "blue") +
  tm_layout(frame = NA)
```

We can use `FedData::get_daymet()` to download Daymet data that covers the spatial extent of the polygons data. The downloaded dataset can be assigned to an R object as `RasterBrick` (you could alternatively write the downloaded data to a file). In order to let the function know the spatial extent for which it download Daymet data, we supply a `SpatialPolygonsDataFrame` object supported by the `sp` package. Since our main vector data handling package is `sf` we need to convert the `sf` object to a `sp` object. 

The code below downloads `prcp` and `tmax` for the spatial extent of Michigan counties for 2000 and 2001:

```{r MI_daymet_download, eval = F}
(
MI_daymet_select <- FedData::get_daymet(
  #--- supply the vector data in sp ---#
  template = as(MI_counties_select, "Spatial"),
  #--- label ---#
  label = "MI_counties_select",
  #--- variables to download ---#
  elements = c("prcp","tmax"),
  #--- years ---#
  years = 2000:2001
)
)

```

```{r MI_daymet_import, echo = F}
# saveRDS(MI_daymet_select, "./Data/tmi_prcp_MI_2000_2001.rds")
(
MI_daymet_select <- readRDS("./Data/tmi_prcp_MI_2000_2001.rds")
)
```

As you can see, Daymet `prcp` and `tmax` data are stored separately as `RasterBrick`. Now that you have stars objects, you can extract values for your target polygons data using the methods described in Chapter \@ref(int-RV). 

If you use the `stars` package for raster data handling (see Chapter \@ref(stars-basics)), you can convert them into `stars` object using `st_as_stars()`.

```{r convert_to_stars}
#--- tmax as stars ---#
tmax_stars <- st_as_stars(MI_daymet_select$tmax)

#--- prcp as stars ---#
prcp_stars <- st_as_stars(MI_daymet_select$prcp)
```

Now, the third dimension (band) is not recognized as dates. We can use `st_set_dimension()` to change that (see Chapter \@ref(stars-set-time)). Before that, we first need to recover `Date` values from the "band" values as follows:

```{r band_to_date}
date_values <- tmax_stars %>% 
  #--- get band values ---#
  st_get_dimension_values(., "band") %>% 
  #--- remove X ---#
  gsub("X", "", .) %>% 
  #--- convert to date ---#
  ymd(.)

#--- take a look ---#
head(date_values)
```

```{r assign_dates}
#--- tmax ---#
st_set_dimensions(tmax_stars, 3, values = date_values, names = "date" )  

#--- prcp ---#
st_set_dimensions(prcp_stars, 3, values = date_values, names = "date" )  
```

Notice that the date dimension has NA for `delta`. This is because Daymet removes observations for December 31 in leap years to make the time dimension 365 consistently across years. This means that there is a one-day gap between "2000-12-30" and "2000-01-01" as you can see below:

```{r one-day-skip}
date_values[364:367]
```



## USGS (under construction) 

Under construction

<!-- [USGS R](https://owi.usgs.gov/R/index.html)

### Groundwater level data

```{r , eval = F}
library(dataRetrieval)
state_ls <- state.abb

data_ne <- whatNWISdata(stateCd = "NE")

KS_gwl <- readNWISdata(
    stateCd="Kansas", 
    startDate = "1980-01-01", 
    endDate = "2020-01-01", 
    service = "gwlevels"
  ) %>% 
  select(site_no, lev_dt, lev_va) %>% 
  rename(date = lev_dt, dwt = lev_va)

KS_site_ls <- KS_gwl[,site_no] %>% unique()

sites_info <- readNWISsite(siteNumbers = KS_site_ls) %>% 
  select(site_no, dec_lat_va, dec_long_va) %>% 
  st_as_sf(coords = c("dec_long_va", "dec_lat_va")) %>% 
  st_set_crs(4269)
```

### Nitrogen concentration

```{r , eval = F}

SaltLake_totalN <- readNWISdata(
  bBox = c(-113.0428, 40.6474, -112.0265, 41.7018), 
  service = "qw", 
  parameterCd = "00600", 
  startDate = "2020-01-01"
)

attributes(SaltLake_totalN)

attr(SaltLake_totalN, "variableInfo")


phosphorous: 00665
nitrogen: 00665
```


### Water temperature
```{r , eval = F}
MauiCo_avgdailyQ <- readNWISdata(
  stateCd = "Hawaii", 
  service = "dv", 
  parameterCd = "00060"
)

head(MauiCo_avgdailyQ)
```

### WQP 

[WQP user guide](https://www.waterqualitydata.us/portal_userguide/)

[WQP query](https://www.waterqualitydata.us/webservices_documentation/#WQPWebServicesGuide-Submitting)

```{r , eval = F}
wqpcounts_ia <- readWQPdata(
  statecode="US:19", 
  querySummary = TRUE
)

IA_lake_phosphorus <- readWQPdata(
  statecode = "IA", 
  siteType = "Lake, Reservoir, Impoundment", 
  characteristicName = "Phosphorus", 
  startDate = "1990-10-01" 
) %>% 
filter(MonitoringLocationIdentifier, ResultMeasureValue, ActivityStartDate)

#--- get longitude and latitude ---#
siteInfo <- attr(IA_lake_phosphorus, "siteInfo") %>% 
  select(MonitoringLocationIdentifier, dec_lat_va, dec_lon_va) %>% 
  st_as_sf(coords = c("dec_lon_va", "dec_lat_va")) %>% 
  st_set_crs(4269)

attributes()
```
 -->

## Sentinel with `sen2r` (under construction)

Under construction


## Census with `tidycensus` (under construction)

Under construction




